\documentclass[12pt,a4paper]{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{url}

\usepackage[authoryear]{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{bookmark}

\geometry{margin=1in}
\setstretch{1.2}

\title{Vesolci obstajajo}
\author{Your Name}
\date{\today}

\begin{document}

% ------------------------
% TITLE PAGE
% ------------------------
\begin{titlepage}
    \centering
    {\large University of Ljubljana\\[0.5cm]
    School of Economics and Business\par}
    \vfill
    {\LARGE \textbf{Portfolio Credit Risk Models:\\A Comprehensive Review}\par}
    \vspace{1.5cm}
    {\large Risk Management 2025/26\\
    Assignment No.\ 2\par}
    \vfill
    {\large
    Author: Alja Dostal, Manca Kavčič\\
    
    Date: December 2025\par}
\end{titlepage}

% ------------------------
% TABLE OF CONTENTS
% ------------------------
\tableofcontents
\newpage

% ------------------------
% ABSTRACT
% ------------------------
\begin{abstract}
This paper provides a comprehensive overview of portfolio credit risk modelling 
frameworks. It presents a taxonomy of models, discusses their  theoretical foundations,
evaluates practical applications, and highlights key strengths and limitations. A
comparative table summarises the major portfolio  credit risk models widely used in practice.
\end{abstract}

% ------------------------
% MAIN TEXT
% ------------------------

\section{Introduction}

Credit risk — the potential loss arising from a counterparty’s failure to meet 
contractual obligations — is one of the central risks managed by financial 
institutions. While traditional credit analysis focuses on individual borrowers, 
portfolio credit risk extends this perspective by explicitly accounting for the 
joint behaviour of multiple obligors, default correlations, and systemic effects. 
Credit risk measurment involves several key components, such as probability of default (PD),
exposure at default (EAD), loss given default (LGD), expected loss (EL) and unexpected loss (UL).
Financial institutions rely on portfolio credit risk models to quantify expected 
loss (EL), unexpected loss (UL), and tail risk measures such as Value-at-Risk (VaR) 
and Expected Shortfall (ES), which are essential for economic capital allocation, 
regulatory compliance, pricing, and portfolio optimisation \citep{hull2018,mcneil2005}.
The paper is based on lecture notes \citep{berk2025}.

\section{Conceptual Foundations of Portfolio Credit Risk}

Portfolio credit losses can be expressed as
\[
L = \sum_{i=1}^{n} EAD_i \cdot LGD_i \cdot D_i,
\]
where $EAD_i$ denotes exposure at default, $LGD_i$ is loss given default, and $D_i$ 
is a default indicator variable. The primary challenge in portfolio credit risk 
modelling lies in capturing the joint distribution of defaults and exposures, as 
losses are rarely independent across obligors \citep{bluhm2016}.

Several fundamental modelling dimensions are commonly used to classify portfolio 
credit risk models. These include the distinction between top-down and bottom-up 
approaches, default-mode versus mark-to-market frameworks, conditional versus 
unconditional models, and structural versus reduced-form specifications. Each 
dimension reflects a different modelling philosophy and has important implications 
for practical applications.

\section{Typology of Portfolio Credit Risk Models}

\subsection{Top-Down and Bottom-Up Models}

Bottom-up models analyse credit risk at the level of individual obligors and then 
aggregate the results to the portfolio level. In this framework, each exposure is 
characterised by its probability of default, exposure at default, and loss given 
default, while portfolio risk emerges from the joint behaviour of defaults across 
obligors. Prominent examples include the Moody's KMV model, CreditMetrics, Gaussian 
copula models, and CreditRisk+ \citep{crouhy2000}. The main advantage of bottom-up 
models lies in their ability to provide detailed and granular risk information, 
which allows for portfolio optimisation, risk attribution, and sensitivity 
analysis at the level of individual borrowers. At the same time, these models 
require extensive obligor-level data and robust estimates of default correlations, 
which can be difficult to obtain in practice. As portfolio size increases, the 
computational burden of simulation-based bottom-up models also becomes significant.

In contrast, top-down models treat the credit portfolio as a single aggregate 
entity and directly model portfolio losses without explicitly modelling individual 
borrowers. Default probabilities are typically conditioned on macroeconomic or 
systematic risk factors, such as economic growth or unemployment rates. Such 
approaches are computationally efficient and are often used for stress testing and 
high-level scenario analysis, particularly when the focus lies on systemic risk 
rather than individual exposures. Their main limitation lies in their lack of 
interpretability and inability to provide borrower-level insights, which restricts 
their usefulness for detailed credit portfolio management \citep{wilson1998}.

\subsection{Default-Mode and Mark-to-Market Models}

Default-mode models focus exclusively on losses that occur at default. In these 
frameworks, credit risk is measured through default probabilities and loss given 
default assumptions, while changes in credit quality prior to default are ignored. 
Expected loss and unexpected loss are therefore driven entirely by the frequency 
and severity of default events. Examples include CreditRisk+, copula-based default 
models, and factor models such as the Vasicek framework \citep{vasicek2002}. These 
models are conceptually simple and computationally efficient, which makes them 
particularly attractive for regulatory capital calculations. However, their 
inability to capture rating migrations or spread dynamics limits their applicability 
for pricing credit-sensitive instruments.

Mark-to-market models extend the analysis by accounting for changes in credit 
quality before default, typically through rating migrations or credit spread 
movements. In this setting, portfolio losses may arise even if no default occurs, 
reflecting deterioration in creditworthiness. CreditMetrics is the most 
well-known example of this approach \citep{jpm1997}. While mark-to-market models 
provide a more comprehensive and economically meaningful view of credit risk, they 
require richer datasets, including rating transition matrices and market-based 
spread information, and involve more complex calibration and simulation procedures.

\subsection{Conditional and Unconditional Default Probability Models}

Probability of default (PB) is the likelihood that over a specified period a borrower will
not be able to make their scheduled repayments on a particular debt. Higher interest
rates are associated with higher default risk.

For businesses, a probability of default is implied by their credit rating from independent rating
agencies, such as S\&P Global Ratings, Fitch Ratings, or Moody's Investors Service, but it can also
be estimated using historical data and statistical techniques. PD is used along with loss given default
(LGD) and exposure at default (EAD) in a variety of risk management models to estimate possible losses
faced by lenders.

The conditional default probability refers to the likelihood that a borrower will default on their
obligations within a specific period, given that they have survived up to that point without defaulting.

\[
\text{Conditional Default Probability} = \frac{\text{Number of Defaults}}{\text{Number of Survivors}}
\]

In contrast, the unconditional default probability reflects the likelihood that a borrower will default
on their obligations within a specific period, irrespective of their survival up to that point. 

\[
\text{Unconditional Default Probability} = \frac{\text{Number of Defaults}}{\text{Total Number of Borrowers}}
\]

The main distinction between conditional and unconditional default probabilities is in the context within which
these probabilities are used. While conditional probabilities focus on survival to a specific point in time,
unconditional probabilities provide a broader assessment of default likelihood. \citep{pyfin2025valuation}

\subsection{Structural and Reduced-Form Models}

Structural credit risk models originate from the seminal work of 
\citep{merton1974}. 
In these models, default occurs when the value of a firm’s 
assets falls below a predefined default barrier, usually related to the book or 
market value of its liabilities. Asset values are assumed to follow a stochastic 
process, and default is an endogenous outcome of the firm’s capital structure and 
asset volatility. Structural models offer strong economic intuition and provide 
market-implied default probabilities that can respond quickly to new information. 
However, they often suffer from unrealistic default timing, as default can occur 
only at specific horizons, and from limited applicability to privately held firms 
for which market data on equity values and volatility are unavailable.

Reduced-form, or intensity-based, models treat default as an exogenous and 
unpredictable jump process governed by a stochastic default intensity 
\citep{jarrow1995}. In this framework, default arrives randomly according to a 
Poisson process, and the probability of default over a given horizon depends on 
the level and dynamics of the default intensity. These models are flexible and 
mathematically tractable and are widely used in the pricing of credit derivatives, 
where the timing of default is crucial. However, reduced-form models lack a direct 
economic interpretation of default events, as default is not explicitly linked to 
the firm’s balance sheet or asset dynamics.

\section{Major Portfolio Credit Risk Models}

\subsection{CreditMetrics}

CreditMetrics, developed by J.P. Morgan in 1997, is a market-to-market credit risk 
model that captures both default and migration risk through rating transition 
matrices, spread volatility, and correlated asset value dynamics \citep{jpm1997}. 
It is a bottom-up model, driven by movements in bonds ratings. Other commitments can
be included, such as bonds, loans, swaps. Impact of market volatility on expected
exposures at maturity has to bo assessed by analysts. Each instrument in the portfolio
is assigned a credit rating, and future rating states are simulated using empirical
transition probabilities. In the event of default, losses are computed using historical
recovery rates differentiated by seniority. Correlations between obligors are introduced
through correlated asset returns, typically assumed to follow a multivariate normal
distribution. The final portfolio loss distribution is obtained via Monte Carlo simulation. The main 
strength of CreditMetrics lies in its ability to integrate credit risk into a 
value-at-risk framework similar to market risk models, while its main weakness 
stems from the difficulty of estimating reliable correlations and transition 
probabilities. \\

\textit{Example.} Consider a single $5$-year corporate bond with face value $F = 10$ million USD and 
current credit rating $r_0 = \text{BBB}$. Let $P_{r_0,r}$ denote the one-year rating 
transition probability from rating $r_0$ to rating $r \in \{A, BBB, BB, D\}$.  

At horizon $T=1$, the mark-to-market value of the bond conditional on rating $r$ is
\[
V_r = \sum_{t=1}^{4} \frac{cF}{(1+s_r)^t} + \frac{F}{(1+s_r)^4},
\]
where $c$ is the coupon rate and $s_r$ is the credit spread associated with rating $r$.
In the event of default ($r=D$), the value is
\[
V_D = (1-\text{LGD})F.
\]

Let asset returns follow a correlated multivariate normal distribution,
\[
\mathbf{X} \sim \mathcal{N}(0,\Sigma),
\]
which determines joint rating migrations across obligors. Monte Carlo simulation 
yields the portfolio value distribution, and the credit value-at-risk at confidence 
level $\alpha$ is
\[
\text{Credit VaR}_\alpha = V_0 - \inf \{ v : \mathbb{P}(V_T \leq v) \geq \alpha \}.
\]

\subsection{CreditRisk+}

CreditRisk+ is an actuarial portfolio credit risk model that treats defaults as 
Poisson-distributed events with sector-specific risk factors \citep{bluhm2016}.
It was published by Credit Suisse in 1997.  
Instead of modelling asset values or credit migrations, the model focuses on the 
frequency and severity of default losses. Default intensity can be modeled with
factors that change in time. Correlation across obligors is introduced 
by grouping exposures into homogeneous sectors that share common systematic risk 
factors. Losses are modelled using exposure bands, and the overall portfolio loss 
distribution is obtained by aggregating losses across bands and sectors. Its 
computational efficiency and modest data requirements make CreditRisk+ attractive 
for large loan portfolios, although it ignores rating migrations and assumes 
simplified dependence structures. \\

\textit{Example.} Consider a portfolio of $N$ homogeneous loans with exposure $E$, loss given default 
$\text{LGD}$, and unconditional default probability $p$. Defaults are modelled as a 
Poisson random variable
\[
D \sim \text{Poisson}(\lambda), \quad \lambda = Np.
\]

Sectoral dependence is introduced via a stochastic default intensity,
\[
\lambda = \sum_{k=1}^{K} w_k \lambda_k,
\]
where $\lambda_k$ are Gamma-distributed sector factors. The total portfolio loss is
\[
L = E \cdot \text{LGD} \cdot D.
\]

The probability generating function of the loss distribution allows analytical 
derivation of expected loss,
\[
\mathbb{E}[L] = E \cdot \text{LGD} \cdot \lambda,
\]
and high-quantile unexpected loss without Monte Carlo simulation.

\subsection{Moody's KMV}

The KMV model builds on Merton’s structural framework and introduces the concept of 
distance-to-default, defined as
\[
DD = \frac{\ln(V/A) + (\mu - \tfrac{1}{2}\sigma^2)T}{\sigma \sqrt{T}}.
\]
This measure captures how far a firm’s asset value is from the default boundary, 
expressed in units of asset volatility.
In Merton's model - which KMV model is based on - value of equity is a call option
on the value of the firm's assets. It is a structural model. The distance-to-default is then mapped to 
empirical default probabilities using large historical databases of default events. 
The model provides forward-looking and market-based default estimates and has been 
shown to anticipate changes in credit ratings. However, it requires high-quality 
market data and relies on simplifying assumptions regarding capital structure, 
asset volatility, and the timing of default 
\citep{merton1974}. \\

\textit{Example.} Let firm asset value $V_t$ follow a geometric Brownian motion,
\[
dV_t = \mu V_t dt + \sigma V_t dW_t.
\]

Define the default point $A$ as short-term debt plus half of long-term debt. The 
distance-to-default at horizon $T$ is
\[
DD = \frac{\ln(V_0/A) + (\mu - \tfrac{1}{2}\sigma^2)T}{\sigma \sqrt{T}}.
\]

The empirical default probability is obtained via a mapping function $f(\cdot)$,
\[
\text{PD} = f(DD),
\]
calibrated from historical default data. This PD can then be used as an input into 
portfolio credit risk models or pricing applications.

\subsection{Credit Portfolio View}

Credit Portfolio View - published by McKinsey in 1997 - explicitly incorporates
macroeconomic variables such as GDP 
growth, unemployment, and interest rates into the modelling of default 
probabilities. Default probabilities are estimated using econometric models, 
typically logit specifications, and vary systematically with the state of the 
economy. Portfolio losses are then simulated conditional on macroeconomic 
scenarios. This approach makes Credit Portfolio View particularly well suited for 
stress testing and scenario analysis. However, its top-down nature limits its 
ability to capture firm-specific credit risk, and its calibration requires long 
and reliable macroeconomic time series \citep{wilson1998}. \\

\textit{Example.} Let the default probability for obligor $i$ at time $t$ be modelled using a logit 
specification,
\[
\log\left(\frac{PD_{i,t}}{1 - PD_{i,t}}\right)
= \beta_0 + \beta_1 \text{GDP}_t + \beta_2 \text{UNEMP}_t.
\]

Given a macroeconomic scenario $\{\text{GDP}_t, \text{UNEMP}_t\}$, conditional default 
probabilities are computed and portfolio loss is simulated as
\[
L_t = \sum_{i=1}^{n} EAD_i \cdot LGD_i \cdot D_{i,t},
\quad D_{i,t} \sim \text{Bernoulli}(PD_{i,t}).
\]

This framework allows scenario-dependent estimation of expected loss and stress 
losses.

\subsection{Gaussian Copula Model}

Gaussian copula model links individual defaults through a common dependence 
structure, typically represented as
\[
D_i = \mathbf{1}(Z_i < \Phi^{-1}(PD_i)).
\]
It models the dependence structure between credit risks of different entities within a portfolio.
The Gaussian copula model operates on the premise that the marginal distribution of each credit
risk is a standard normal distribution. The model then uses a copula function to describe the
correlation structure between these standardized credit risks. By doing so, it allows us to simulate
joint defaults and evaluate portfolio risk. 
In this framework, default occurs when a latent variable falls below a threshold 
determined by the obligor’s marginal default probability. Dependence across 
obligors is captured through correlations among the latent variables. These models 
gained widespread popularity in structured credit markets due to their analytical 
convenience and flexibility. However, one of the most notable limitations is its assumption of normality,
which might not hold in all cases. In situations of severe market stress or extreme events, the model’s
accuracy can be compromised \citep{pyfin2025valuation}. 
%their inability to capture extreme tail 
%dependence became evident during the global financial crisis, leading to systematic 
%underestimation of joint default risk \citep{crouhy2000}.

\subsection{Vasicek One-Factor Model}

The Vasicek one-factor model, which forms the basis of the Basel II and Basel III 
IRB capital framework, represents default risk as driven by a single systematic 
factor: %\citep{gordy2003}
\[
Z_i = \sqrt{\rho} Y + \sqrt{1-\rho} \epsilon_i.
\]
The model offers a structured approach to quantifying default rates by taking into account the
influence of economic factors. It operates within the realm of one-factor models, where a single
systematic factor plays a pivotal role in driving credit events. This factor is typically associated
with macroeconomic conditions, capturing the broad economic environment that affects borrowers’
ability to meet their obligations \citep{pyfin2025valuation}.
% In this framework, default correlations arise naturally from exposure to the common 
% systematic factor. The model offers closed-form solutions for portfolio loss 
% quantiles and provides a transparent regulatory standard. However, the assumption 
% of a single risk factor oversimplifies correlation structures, particularly during 
% periods of financial stress.

\section{Comparative Summary of Models}

% \begin{table}[h!]
% \centering
% \begin{tabular}{|p{1.5cm}|p{2cm}|p{1.3cm}|p{2cm}|p{3cm}|p{3cm}|}
% \hline
% \textbf{Model} & \textbf{Type} & \textbf{Mode} & \textbf{Inputs} & \textbf{Strengths} & \textbf{Weaknesses} \\
% \hline
% Credit Metrics & Bottom-up & MTM & Transition matrices, correlations & Captures migrations; detailed VaR & Correlation estimation; rating lag \\
% \hline
% KMV/ Merton & Structural & MTM & Equity, volatility, liabilities & Economic foundation; market PDs & Requires market data; barrier issues \\
% \hline
% Credit Risk+ & Bottom-up & Default & PDs, exposures, sector factors & Efficient; minimal data & No migration; Poisson assumption \\
% \hline
% Gaussian Copula & Bottom-up & Default & PDs, correlations & Flexible dependence modelling & Underestimates tail risk \\
% \hline
% Credit Portfolio View & Conditional & MTM & Macro economic variables & Ideal for stress tests & Calibration complexity \\
% \hline
% Vasicek (Basel) & Bottom-up & Default & PD, LGD, $\rho$ & Closed forms; regulatory standard & One-factor oversimplifies correlation \\
% \hline
% \end{tabular}
% \caption{Comparative summary of major portfolio credit risk models.}
% \end{table}

%\section{Comparative Summary of Portfolio Credit Risk Models}

\begin{table}[h!]
\centering
\small
\begin{tabular}{|p{2.5cm}|p{3.5cm}|p{3.5cm}|p{4cm}|}
\hline
\textbf{Model} 
& \textbf{Type} 
& \textbf{Key Inputs}
& \textbf{Key Remarks} \\
\hline

CreditMetrics 
& Bottom-up, MTM, Conditional, Structural 
& Rating transitions, correlations, spreads 
& Explicit migration risk; MTM valuation sensitive to spread dynamics \\
\hline

CreditRisk+ 
& Bottom-up, Default, Unconditional, Reduced-form 
& PDs, exposures, sector factors 
& Defaults follow Poisson law; no migration or MTM effects \\
\hline

Moody's KMV
& Bottom-up, MTM, Conditional, Structural 
& Equity value, volatility, debt structure 
& Defaults linked to firm value crossing barrier; market-implied PDs \\
\hline

Credit Portfolio View 
& Top-down, MTM, Conditional, Reduced-form 
& Macroeconomic variables, credit spreads 
& PDs conditioned on macro states; ideal for stress testing \\
\hline

Gaussian Copula 
& Bottom-up, Default, Unconditional, Reduced-form 
& Marginal PDs, asset correlations 
& Standard dependence model for CDOs; weak tail dependence \\
\hline


Vasicek (Basel ASRF) 
& Bottom-up, Default, Conditional (from unconditional PD), Reduced-form 
& PD, LGD, asset correlation $\rho$ 
& One-factor model; regulatory benchmark for capital \\
\hline

% Factor Copula Models 
% & Top-down 
% & Default 
% & Conditional 
% & Reduced-form 
% & Latent factors, PD term structures 
% & Generalisation of Gaussian copula; improved tail modelling \\
% \hline

\end{tabular}
\caption{Comparative summary of major portfolio credit risk models.}
\end{table}

% \section{Discussion}

% Portfolio credit risk modelling has evolved from economically intuitive structural 
% models to actuarial and copula-based frameworks, as well as regulatory factor 
% models. Following the global financial crisis, greater emphasis has been placed 
% on stress testing and macroeconomic sensitivity. In practice, institutions often 
% combine multiple models to compensate for individual limitations.

%\newpage

\section{Conclusion}

% Portfolio credit risk models are indispensable tools in modern risk management. 
% Each modelling framework embodies specific assumptions and trade-offs, making it 
% essential to understand their limitations and appropriate use cases. Future 
% developments are likely to further integrate macro-financial feedback mechanisms 
% and advanced analytical techniques.

Portfolio credit risk modelling seeks to quantify losses arising from correlated
default events in a manner that is both theoretically sound and practically implementable.
This paper has shown that no single modelling framework can fully capture the complexity
of credit risk, as each approach reflects different assumptions about default behaviour,
dependence structures, data availability, and intended use cases.

Bottom-up models offer granular insights and are well suited for portfolio management and
regulatory capital assessment, while top-down approaches provide computationally efficient
tools for stress testing and macroeconomic scenario analysis. Similarly, default-mode models
prioritise simplicity and robustness, making them attractive for large portfolios and regulatory
purposes, whereas mark-to-market models deliver a richer representation of credit risk by
incorporating migration and spread dynamics. Structural models ground default events in firm
value dynamics and offer strong economic intuition, but reduced-form models provide greater
flexibility and are better suited for modelling default timing and pricing credit derivatives.

This paper shows that each of the major portfolio credit risk models - CreditMetrics, CreditRisk+,
Moody's KMV, Credit Portfolio View, Gaussian copula model, and the Vasicek one-factor framework -
is not a universal solution and is rather used as a specialised tool. It is best to use multiple
models in complementary ways to address different dimensions of credit risk.

Ultimately, effective portfolio credit risk management depends not only on model sophistication but
also on sound judgement, high-quality data, and a clear understanding of model limitations. Recognising
these limitations is essential, particularly in stressed market environments where simplifying assumptions
may break down.

% ------------------------
% BIBLIOGRAPHY
% ------------------------

%\newpage

\nocite*{}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
